Paper 1: Triangulating trends in inequality around the world from secondary databases

### 2023-05-03 PhD_pages/Paper_1/data_appendix/1_A_prepare_PIP.qmd
Paper 1: restructuring to prep data separate to paper

import pandas as pd

pd.set_option('display.max_rows', None)
```
:::

::: {.callout-note collapse="true" appearance="minimal"}
### Load PIP data
We load the data from a file prepared from the PIP API by Our World in Data.

```{python}
# df_pip<- read.csv("https://raw.githubusercontent.com/owid/poverty-data/main/datasets/pip_dataset.csv")

@@

@@

@@

keep_cols = [
  "Year",
  "Entity",
  "Gini",
  "Top 10pc share",
  "Bottom 50pc share",
  "MLD",
  "welfare_type"
]

>>>>

import pandas as pd
import json
import requests

::: {.callout-note collapse="true" appearance="minimal"}
### Load PIP percentile data
```{python}
fp = "data/original/pip_percentiles.csv"

df_pip_percentiles = pd.read_csv(fp)

::: {.callout-note collapse="true" appearance="minimal"}
### Keep top 1% shares
```{python}

drop_cols = ['percentile', 'avg_welfare', 'pop_share', 'quantile']

df_pip_percentiles = df_pip_percentiles[df_pip_percentiles['percentile'] == 100]\
  .drop(drop_cols, axis=1)\
  .rename(columns={
    'welfare_share': 'Top 1pc share'
  })

# Convert shares to percent
df_pip_percentiles['Top 1pc share'] = df_pip_percentiles['Top 1pc share'] * 100
```

::: {.callout-note collapse="true" appearance="minimal"}
### Standardize country names in percentile data

```{python}
# # Initially I save a csv of country-codes included in the percentil data, to pass through the OWID country name standardization tool
# unique_countries = df_pip_percentiles['country_code'].unique()
# df_pip_percentiles_countries = pd.DataFrame({'country': unique_countries})

# df_pip_percentiles_countries.to_csv('data/original/pip_percentiles_countries.csv', index=False)

# Read in name mapping file
fp = 'data/original/pip_percentiles_countries_country_standardized.csv'
name_mapping = pd.read_csv(fp)

# Merge into

# Merge mapping into data

df_pip_percentiles = pd.merge(df_pip_percentiles, name_mapping, left_on='country_code', right_on='country', how = 'left')

# Drop columns used in the construction of the standardized country column
df_pip_percentiles = df_pip_percentiles.drop(columns=['country','country_code'])

df_pip_percentiles = df_pip_percentiles.rename(columns = {'Our World In Data Name': 'country'})
```

```{python}
df_pip_percentiles.head()
```
:::

::: {.callout-note collapse="true" appearance="minimal"}
### Load PIP main data (already prepared by OWID)
We load the data from a file prepared from the PIP API by Our World in Data.

```{python}
# df_pip = pd.read_csv("https://raw.githubusercontent.com/owid/poverty-data/main/datasets/pip_dataset.csv")

@@

::: {.callout-note collapse="true" appearance="minimal"}
### Merge top 1% shares from percentile data into main PIP data

```{python}
df_pip = pd.merge(df_pip, df_pip_percentiles, on=['country', 'year', 'reporting_level', 'welfare_type'], how='left')

df_pip.head()
```
:::

@@

# Top 10 / bottom 50% share
df_pip["Ratio Top 10 Bottom 50 share"] = df_pip["decile10_share"]/df_pip["Bottom 50pc share"] * 100

@@

keep_cols = [
  "Year",
  "Entity",
  "Gini",
  "Top 1pc share",
  "Top 10pc share",
  "Bottom 50pc share",
  "Ratio Top 10 Bottom 50 share",
  "MLD",
  "welfare_type"
]

### 2023-05-03 PhD_pages/Paper_1/paper1.qmd
Paper 1: restructuring to prep data separate to paper

# This specifies the +/- for the 'stable' category in the tables below
tolerance_lookup =  {
                'var_search_term': [
                    'Gini',
                    'Top 1pc share', 
                    'Top 10pc share',
                    'Bottom 50pc share'],
                'var_tolerance': [
                    1,
                    1,
                    1,
                    1
                ], 
                'relative_tolerance': [
                    0.05
                ]
            }

@@

@@

Summary, across both periods. (Reflect this back in the intro.)

@@

@@

@@

# Bridging the gap for particular countries / welfare concept

"Korinek, Mistiaen, and
Ravallion (2006) estimate that correcting
for such selective compliance adds about
five percentage points to the Gini index for
the United States."

## Sub-Saharan Africa
Notes on the WID methodology. Basically no tax data. Based on guesses about how you shift consumption to pre-tax income, and then top income adjustment.

Plot: make a direct comparison for this region with the same index. NEcessarily tethered to the survey data (unlike say Mexico) The changes are amplified in WID (falls are larger, rises are larger).

Also note the quality/comparability of the PIP survey data.

-----

### Aligning coverage

## Global data on inequality

### Global databases

A global view faces the hard limitation of the underlying data. This has two sides: firstly is there any data? Secondly the comparability of the data, across countries and over time.

Much improved – *harmonized* secondary databases. Three of note.

- PIP

- WID

- LIS

### Two worlds of inequality measurement

#### Connecting up the worlds is valuable

- for Policy
- for academic research

## Triangulating a global view

### Since 1990

### Since 1980

### Comparing metrics

### Since 1990

- India: Insane rise in WID 34->56
- China:
- Russia: Massive drop in PIP. Insane rise in WID.
- USA: Much larger rise in WID.

What explains the different trends in these countries? (We can maybe look in more detail, at different welfare concepts etc.)

It sometimes goes the other way: Indonesia – Rise in PIP, fall in WID.

:::

#### Method 2: Use a regression to obtain the time trend

### Results for a range of inequality metrics

::: {.callout-note collapse="true" appearance="minimal"} 
## Calculations for comparisons

## Explaining differences between PIP and WID data

### For a subset of countries we can bridge between the PIP and WID data

Besides the underlying data sources, there are many differences between what is being measured in the PIP and WID data - notably the welfare concept and unit of analysis/equivalization.

In order to get a better understanding of where differences between the datasets arise, for a subset of countries we can bridge between the two somewhat. For example:

-   WID has post-tax income series for some countries
-   We can use LIS to look at how changing the welfare concept, unit of analysis, equivalisation affects things. (For equivalisation, we could also possibly use the newly available PIP harmonized microdata).

### More in-depth discussion of particular outliers

-   Russia is a very notable case where the two datasets tell very different stories.

>>>>

# This specifies the +/- for the 'stable' category in the tables below
tolerance_lookup =  {
                'var_search_term': [
                    'Gini',
                    'Top 1pc share', 
                    'Top 10pc share',
                    'Bottom 50pc share',
                    "Ratio Top 10 Bottom 50 share"],
                'var_tolerance': [
                    1,
                    1,
                    1,
                    1,
                    1
                ], 
                'relative_tolerance': [
                    0.05
                ]
            }

@@

- Plot forthe appendix: make a direct comparison for Africa region with the same inequality index. Necessarily tethered to the survey data (unlike say Mexico) The changes are amplified in WID (falls are larger, rises are larger).

@@

@@

- India: Insane rise in WID 34->56
- Russia: Massive drop in PIP. Insane rise in WID.
- USA: Much larger rise in WID.

It sometimes goes the other way: 
- Indonesia: Rise in PIP, fall in WID.
- Some SSA countries: Larger falls in WID.

@@

- For a range of metrics, check which countries disagree on change category using a relative definitnio (+/-5%). And report global averages.

```{python}
# Set-up to compare PIP
reference_vals = [1993, 2015]
max_dist_from_refs = 3

selected_vars = [
        'PIP: Gini',
        'PIP: Ratio Top 10 Bottom 50 share'
        ]

output_dict_PIP_indices_1993 = prep_data_and_change_summary(
      source_dfs = source_dfs,
      df_pop = df_pop,
      df_pop_regions = df_pop_regions,
      df_regions = df_regions,
      region_col = 'region',
      selected_vars = selected_vars,
      reference_vals = reference_vals,
      max_dist_from_refs = max_dist_from_refs,
      min_dist_between = 1,
      reference_col = "Year",
      group_by_col = "Entity",
      outlier_cut_off_upper = None,
      tolerance_lookup =  tolerance_lookup
      )

```{python}
#| label: tbl-pip-indices-counts
#| tbl-cap: Change in inequality 1980 vs 2018 - All availalbe observations
output_dict_PIP_indices_1993['all_sources_False']['absolute_counts']

```{python}
#| label: tbl-pip-indices-avgs
#| tbl-cap: Unweighted and weighted average in 1980 vs 2018 - All availalbe observations
output_dict_PIP_indices_1993['all_sources_False']['averages']

@@

# 5. Bridging the gap for particular countries / welfare concept
